### 2024-08-15 machine-learning.qmd
Built site for gh-pages

Machine learning can be useful for leveraging big data and lots of predictor variable to develop predictive models with greater accuracy.
However, many machine learning techniques are black boxes—it is often unclear
Machine learning tends to be a data-driven, atheoretical technique.
This can result in overfitting.
Thus, when estimating machine learning models, it is common to keep a hold-out sample for use in cross-validation to evaluate the extent of shrinkage of model coefficients.
The data that the model is trained on is known as the "training data".
The data that the model was not trained on but is then is independently tested on—i.e., the hold-out sample—is the "test data".
Shrinkage occurs when predictor variables explain some random error variance in the original model.
When the model is applied to an independent sample (i.e., the test data), the predictive model will likely not perform quite as well, and the regressions coefficients will tend to get smaller (i.e., shrink).

@@

### Fill in Missing Data {#sec-fillMissingData}

@@

```{r}
seasonalData_lag_subsetQB <- seasonalData_lag_subset %>% 
  filter(position == "QB") %>% 
  select(
    gsis_id, season, games, gs, years_of_experience, age, ageCentered20, ageCentered20Quadratic,
    height, weight, rookie_year, draft_number,
    fantasy_points, fantasy_points_ppr, fantasyPoints_lag,
    completions:rushing_2pt_conversions, contains(".pass"))

@@

@@

# QBs
seasonalData_lag_subsetQB_imp <- missForest::missForest(
  seasonalData_lag_subsetQB %>% 
    select(-gsis_id) %>%
    data.frame(),
  verbose = TRUE,
  mtry = 100,
  parallelize = "forests")

seasonalData_lag_subsetQB_imp$OOBerror
data_train_qb <- cbind(seasonalData_lag_subsetQB[,"gsis_id"], seasonalData_lag_subsetQB_imp$ximp)
data_train_qb_matrix <- data_train_qb %>%
  mutate(across(where(is.factor), ~ as.numeric(as.integer(.)))) %>% 
  as.matrix()

newData_seasonalQB_imp <- missForest::missForest(
  newData_seasonalQB %>% 
    select(-gsis_id) %>%
    data.frame(),
  verbose = TRUE,
  mtry = 100,
  parallelize = "forests")

newData_seasonalQB_imp$OOBerror
newData_qb <- cbind(newData_seasonalQB[,"gsis_id"], newData_seasonalQB_imp$ximp)
newData_qb_matrix <- newData_qb %>%
  mutate(across(where(is.factor), ~ as.numeric(as.integer(.)))) %>% 
  as.matrix()

# RBs
seasonalData_lag_subsetRB_imp <- missForest::missForest(
  seasonalData_lag_subsetRB %>% 
    select(-gsis_id) %>%
    data.frame(),
  verbose = TRUE,
  mtry = 100,
  parallelize = "forests")

seasonalData_lag_subsetRB_imp$OOBerror

newData_seasonalRB_imp <- missForest::missForest(
  newData_seasonalRB %>% 
    select(-gsis_id) %>%
    data.frame(),
  verbose = TRUE,
  mtry = 100,
  parallelize = "forests")

newData_seasonalRB_imp$OOBerror

# WRs
seasonalData_lag_subsetWR_imp <- missForest::missForest(
  seasonalData_lag_subsetWR %>% 
    select(-gsis_id) %>%
    data.frame(),
  verbose = TRUE,
  mtry = 100,
  parallelize = "forests")

seasonalData_lag_subsetWR_imp$OOBerror

newData_seasonalWR_imp <- missForest::missForest(
  newData_seasonalWR %>% 
    select(-gsis_id) %>%
    data.frame(),
  verbose = TRUE,
  mtry = 100,
  parallelize = "forests")

newData_seasonalWR_imp$OOBerror

# TEs
seasonalData_lag_subsetTE_imp <- missForest::missForest(
  seasonalData_lag_subsetTE %>% 
    select(-gsis_id) %>%
    data.frame(),
  verbose = TRUE,
  mtry = 100,
  parallelize = "forests")

seasonalData_lag_subsetTE_imp$OOBerror

newData_seasonalTE_imp <- missForest::missForest(
  newData_seasonalTE %>% 
    select(-gsis_id) %>%
    data.frame(),
  verbose = TRUE,
  mtry = 100,
  parallelize = "forests")

newData_seasonalTE_imp$OOBerror

@@

#pred_vars_qb_categorical <- ""

@@

#gpb.Dataset.set.categorical(gp_data_qb, pred_vars_qb_categorical)

params_qb <- list(
  learning_rate = 0.01,
  max_depth = 2,
  num_leaves = 2^10,
  min_data_in_leaf = 10,
  lambda_l2 = 10,
  num_threads = num_cores)
  #optimizer_cov = "nelder_mead")

@@

gp_model_qb$set_optim_params(params = list(optimizer_cov = "nelder_mead"))
summary(gp_model_qb) # Estimated random effects model

gp_model_fit_qb <- gpb.train(
  data = gp_data_qb,
  gp_model = gp_model_qb,
  nrounds = nrounds_qb,
  params = params_qb,
  verbose = 0)

# Model predictions
pred_qb <- predict(
  gp_model_fit_qb,
  data = newData_qb_matrix[,pred_vars_qb],
  group_data_pred = newData_qb_matrix[,"gsis_id"],
  group_rand_coef_data_pred = cbind(
    newData_qb_matrix[,"ageCentered20"],
    newData_qb_matrix[,"ageCentered20Quadratic"]),
  predict_var = FALSE,
  pred_latent = FALSE)

y_pred_qb <- pred_qb[["response_mean"]]
mean((newData_qb[,"fantasyPoints_lag"] - y_pred_qb)^2)
```

>>>>

Machine learning takes us away from focusing on [causal inference](#sec-causalInference).
Machine learning does not care about which processes are causal—i.e., which processes influence the outcome.
Instead, machine learning cares about prediction—it cares about a predictor variable to the extent that it increases predictive accuracy regardless of whether it is causally related to the outcome.

Machine learning can be useful for leveraging big data and lots of predictor variable to develop predictive models with greater accuracy.
However, many machine learning techniques are black boxes—it is often unclear how or why certain predictions are made, which can make it difficult to interpret the model's decisions and understand the underlying relationships between variables.
Machine learning tends to be a data-driven, atheoretical technique.
This can result in overfitting.
Thus, when estimating machine learning models, it is common to keep a hold-out sample for use in cross-validation to evaluate the extent of shrinkage of model coefficients.
The data that the model is trained on is known as the "training data".
The data that the model was not trained on but is then is independently tested on—i.e., the hold-out sample—is the "test data".
Shrinkage occurs when predictor variables explain some random error variance in the original model.
When the model is applied to an independent sample (i.e., the test data), the predictive model will likely not perform quite as well, and the regressions coefficients will tend to get smaller (i.e., shrink).

@@

### Fill in Missing Data for Static Variables {#sec-fillMissingData}

@@

```{r}
seasonalData_lag_subsetQB <- seasonalData_lag_subset %>% 
  filter(position == "QB") %>% 
  select(
    gsis_id, season, games, gs, years_of_experience, age, ageCentered20, ageCentered20Quadratic,
    height, weight, rookie_year, draft_number,
    fantasy_points, fantasy_points_ppr, fantasyPoints_lag,
    completions:rushing_2pt_conversions, special_teams_tds, contains(".pass"), contains(".rush"))

@@

::: {#nte-machineLearningImputeMissingData .callout-note title="Impute missing data for machine learning"}
Note: the following code takes a while to run.
:::

@@

# QBs
seasonalData_lag_qb_all_imp <- missForest::missForest(
  seasonalData_lag_qb_all %>% 
    select(-gsis_id) %>%
    data.frame(),
  verbose = TRUE,
  mtry = 100,
  parallelize = "forests")

seasonalData_lag_qb_all_imp$OOBerror
data_all_qb <- cbind(seasonalData_lag_qb_all[,"gsis_id"], seasonalData_lag_qb_all_imp$ximp)
data_all_qb_matrix <- data_all_qb %>%
  mutate(across(where(is.factor), ~ as.numeric(as.integer(.)))) %>% 
  as.matrix()
newData_qb <- data_all_qb %>% 
  filter(season == max(season, na.rm = TRUE)) %>% 
  select(-fantasyPoints_lag)
newData_qb_matrix <- data_all_qb_matrix[
  data_all_qb_matrix[, "season"] == max(data_all_qb_matrix[, "season"], na.rm = TRUE), # keep only rows with the most recent season
  , # all columns
  drop = FALSE]

dropCol_qb <- which(colnames(newData_qb_matrix) == "fantasyPoints_lag")
newData_qb_matrix <- newData_qb_matrix[, -dropCol_qb, drop = FALSE]

seasonalData_lag_qb_train_imp <- missForest::missForest(
  seasonalData_lag_qb_train %>% 
    select(-gsis_id) %>%
    data.frame(),
  verbose = TRUE,
  mtry = 100,
  parallelize = "forests")

seasonalData_lag_qb_train_imp$OOBerror
data_train_qb <- cbind(seasonalData_lag_qb_train[,"gsis_id"], seasonalData_lag_qb_train_imp$ximp)
data_train_qb_matrix <- data_train_qb %>%
  mutate(across(where(is.factor), ~ as.numeric(as.integer(.)))) %>% 
  as.matrix()

seasonalData_lag_qb_test_imp <- missForest::missForest(
  seasonalData_lag_qb_test %>% 
    select(-gsis_id) %>%
    data.frame(),
  verbose = TRUE,
  mtry = 100,
  parallelize = "forests")

seasonalData_lag_qb_test_imp$OOBerror
data_test_qb <- cbind(seasonalData_lag_qb_test[,"gsis_id"], seasonalData_lag_qb_test_imp$ximp)
data_test_qb_matrix <- data_test_qb %>%
  mutate(across(where(is.factor), ~ as.numeric(as.integer(.)))) %>% 
  as.matrix()

# RBs
seasonalData_lag_rb_all_imp <- missForest::missForest(
  seasonalData_lag_rb_all %>% 
    select(-gsis_id) %>%
    data.frame(),
  verbose = TRUE,
  mtry = 100,
  parallelize = "forests")

seasonalData_lag_rb_all_imp$OOBerror
data_all_rb <- cbind(seasonalData_lag_rb_all[,"gsis_id"], seasonalData_lag_rb_all_imp$ximp)
data_all_rb_matrix <- data_all_rb %>%
  mutate(across(where(is.factor), ~ as.numeric(as.integer(.)))) %>% 
  as.matrix()
newData_rb <- data_all_rb %>% 
  filter(season == max(season, na.rm = TRUE)) %>% 
  select(-fantasyPoints_lag)
newData_rb_matrix <- data_all_rb_matrix[
  data_all_rb_matrix[, "season"] == max(data_all_rb_matrix[, "season"], na.rm = TRUE), # keep only rows with the most recent season
  , # all columns
  drop = FALSE]

dropCol_rb <- which(colnames(newData_rb_matrix) == "fantasyPoints_lag")
newData_rb_matrix <- newData_rb_matrix[, -dropCol_rb, drop = FALSE]

seasonalData_lag_rb_train_imp <- missForest::missForest(
  seasonalData_lag_rb_train %>% 
    select(-gsis_id) %>%
    data.frame(),
  verbose = TRUE,
  mtry = 100,
  parallelize = "forests")

seasonalData_lag_rb_train_imp$OOBerror
data_train_rb <- cbind(seasonalData_lag_rb_train[,"gsis_id"], seasonalData_lag_rb_train_imp$ximp)
data_train_rb_matrix <- data_train_rb %>%
  mutate(across(where(is.factor), ~ as.numeric(as.integer(.)))) %>% 
  as.matrix()

seasonalData_lag_rb_test_imp <- missForest::missForest(
  seasonalData_lag_rb_test %>% 
    select(-gsis_id) %>%
    data.frame(),
  verbose = TRUE,
  mtry = 100,
  parallelize = "forests")

seasonalData_lag_rb_test_imp$OOBerror
data_test_rb <- cbind(seasonalData_lag_rb_test[,"gsis_id"], seasonalData_lag_rb_test_imp$ximp)
data_test_rb_matrix <- data_test_rb %>%
  mutate(across(where(is.factor), ~ as.numeric(as.integer(.)))) %>% 
  as.matrix()

# WRs
seasonalData_lag_wr_all_imp <- missForest::missForest(
  seasonalData_lag_wr_all %>% 
    select(-gsis_id) %>%
    data.frame(),
  verbose = TRUE,
  mtry = 100,
  parallelize = "forests")

seasonalData_lag_wr_all_imp$OOBerror
data_all_wr <- cbind(seasonalData_lag_wr_all[,"gsis_id"], seasonalData_lag_wr_all_imp$ximp)
data_all_wr_matrix <- data_all_wr %>%
  mutate(across(where(is.factor), ~ as.numeric(as.integer(.)))) %>% 
  as.matrix()
newData_wr <- data_all_wr %>% 
  filter(season == max(season, na.rm = TRUE)) %>% 
  select(-fantasyPoints_lag)
newData_wr_matrix <- data_all_wr_matrix[
  data_all_wr_matrix[, "season"] == max(data_all_wr_matrix[, "season"], na.rm = TRUE), # keep only rows with the most recent season
  , # all columns
  drop = FALSE]

dropCol_wr <- which(colnames(newData_wr_matrix) == "fantasyPoints_lag")
newData_wr_matrix <- newData_wr_matrix[, -dropCol_wr, drop = FALSE]

seasonalData_lag_wr_train_imp <- missForest::missForest(
  seasonalData_lag_wr_train %>% 
    select(-gsis_id) %>%
    data.frame(),
  verbose = TRUE,
  mtry = 100,
  parallelize = "forests")

seasonalData_lag_wr_train_imp$OOBerror
data_train_wr <- cbind(seasonalData_lag_wr_train[,"gsis_id"], seasonalData_lag_wr_train_imp$ximp)
data_train_wr_matrix <- data_train_wr %>%
  mutate(across(where(is.factor), ~ as.numeric(as.integer(.)))) %>% 
  as.matrix()

seasonalData_lag_wr_test_imp <- missForest::missForest(
  seasonalData_lag_wr_test %>% 
    select(-gsis_id) %>%
    data.frame(),
  verbose = TRUE,
  mtry = 100,
  parallelize = "forests")

seasonalData_lag_wr_test_imp$OOBerror
data_test_wr <- cbind(seasonalData_lag_wr_test[,"gsis_id"], seasonalData_lag_wr_test_imp$ximp)
data_test_wr_matrix <- data_test_wr %>%
  mutate(across(where(is.factor), ~ as.numeric(as.integer(.)))) %>% 
  as.matrix()

# TEs
seasonalData_lag_te_all_imp <- missForest::missForest(
  seasonalData_lag_te_all %>% 
    select(-gsis_id) %>%
    data.frame(),
  verbose = TRUE,
  mtry = 100,
  parallelize = "forests")

seasonalData_lag_te_all_imp$OOBerror
data_all_te <- cbind(seasonalData_lag_te_all[,"gsis_id"], seasonalData_lag_te_all_imp$ximp)
data_all_te_matrix <- data_all_te %>%
  mutate(across(where(is.factor), ~ as.numeric(as.integer(.)))) %>% 
  as.matrix()
newData_te <- data_all_te %>% 
  filter(season == max(season, na.rm = TRUE)) %>% 
  select(-fantasyPoints_lag)
newData_te_matrix <- data_all_te_matrix[
  data_all_te_matrix[, "season"] == max(data_all_te_matrix[, "season"], na.rm = TRUE), # keep only rows with the most recent season
  , # all columns
  drop = FALSE]

dropCol_te <- which(colnames(newData_te_matrix) == "fantasyPoints_lag")
newData_te_matrix <- newData_te_matrix[, -dropCol_te, drop = FALSE]

seasonalData_lag_te_train_imp <- missForest::missForest(
  seasonalData_lag_te_train %>% 
    select(-gsis_id) %>%
    data.frame(),
  verbose = TRUE,
  mtry = 100,
  parallelize = "forests")

seasonalData_lag_te_train_imp$OOBerror
data_train_te <- cbind(seasonalData_lag_te_train[,"gsis_id"], seasonalData_lag_te_train_imp$ximp)
data_train_te_matrix <- data_train_te %>%
  mutate(across(where(is.factor), ~ as.numeric(as.integer(.)))) %>% 
  as.matrix()

seasonalData_lag_te_test_imp <- missForest::missForest(
  seasonalData_lag_te_test %>% 
    select(-gsis_id) %>%
    data.frame(),
  verbose = TRUE,
  mtry = 100,
  parallelize = "forests")

seasonalData_lag_te_test_imp$OOBerror
data_test_te <- cbind(seasonalData_lag_te_test[,"gsis_id"], seasonalData_lag_te_test_imp$ximp)
data_test_te_matrix <- data_test_te %>%
  mutate(across(where(is.factor), ~ as.numeric(as.integer(.)))) %>% 
  as.matrix()

@@

#pred_vars_qb_categorical <- "" # to specify categorical predictors

@@

#gpb.Dataset.set.categorical(gp_data_qb, pred_vars_qb_categorical) # to specify categorical predictors

params_qb <- list(
  learning_rate = 0.01,
  max_depth = 2,
  num_leaves = 2^10,
  min_data_in_leaf = 10,
  lambda_l2 = 10,
  num_threads = num_cores)

@@

gp_model_qb$set_optim_params(params = list(optimizer_cov = "nelder_mead"))

gp_model_fit_qb <- gpb.train(
  data = gp_data_qb,
  gp_model = gp_model_qb,
  nrounds = nrounds_qb,
  params = params_qb) # verbose = 0

summary(gp_model_qb) # Estimated random effects model

# Test Model on Test Data
pred_test_qb <- predict(
  gp_model_fit_qb,
  data = data_test_qb_matrix[,pred_vars_qb],
  group_data_pred = data_test_qb_matrix[,"gsis_id"],
  group_rand_coef_data_pred = cbind(
    data_test_qb_matrix[,"ageCentered20"],
    data_test_qb_matrix[,"ageCentered20Quadratic"]),
  predict_var = FALSE,
  pred_latent = FALSE)

y_pred_test_qb <- pred_test_qb[["response_mean"]]
mean((data_test_qb_matrix[,"fantasyPoints_lag"] - y_pred_test_qb)^2)

# Generate model predictions for next season
pred_nextYear_qb <- predict(
  gp_model_fit_qb,
  data = newData_qb_matrix[,pred_vars_qb],
  group_data_pred = newData_qb_matrix[,"gsis_id"],
  group_rand_coef_data_pred = cbind(
    newData_qb_matrix[,"ageCentered20"],
    newData_qb_matrix[,"ageCentered20Quadratic"]),
  predict_var = FALSE,
  pred_latent = FALSE)

newData_qb$fantasyPoints_lag <- pred_nextYear_qb$response_mean

# Merge with player names
newData_qb <- left_join(
  newData_qb,
  nfl_playerIDs %>% select(gsis_id, name),
  by = "gsis_id"
)

newData_qb %>% 
  arrange(-fantasyPoints_lag) %>% 
  select(name, fantasyPoints_lag)
```
